---
title       : NBA Individual Statistics App
subtitle    : 
author      : Dimitri Aline
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## Two steps

# 1. Choose Player

 Easy to choose a player by selecting their team and position.
 
# 2. Show individual Stats

 Show Complete Stats in a table

 Illustrate stats with graphics
  1. General Stats
  2. Shooting Stats

--- 

## Choose Player

The app uses the following function named listplayer.
This function give the complete list of the players with a given Team and a given Position. 
The Team and Position are the two arguments of this function.
```{r,echo=FALSE}
library("SportsAnalytics")
NBA <- fetch_NBAPlayerStatistics(season = "13-14")
NBA$Team <- as.vector(NBA$Team)
NBA$Position <- as.vector(NBA$Position)
```
```{r,echo=TRUE}
listplayer <- function(team,position){
  team <- as.character(team)
  position <- as.character(position)
  player <- NBA[(NBA$Team == team) & (NBA$Position == position),]  
  player <- player$Name  
  player  
}
```
Here, you can see an exemple with the team Boston and the position Point Guard.
```{r,echo=TRUE}
listplayer("BOS","PG")
```

---

## Show individual Stats

Here you can see an exemple of the general stat plot with the player Rajon Rondo.

```{r,echo=FALSE}
library(ggplot2)
NBA[,6] <- NBA[,6]/NBA[,5]
NBA[,7] <- NBA[,7]/NBA[,5]
NBA[,8] <- NBA[,8]/NBA[,5]
NBA[,9] <- NBA[,9]/NBA[,5]
NBA[,10] <- NBA[,10]/NBA[,5]
NBA[,11] <- NBA[,11]/NBA[,5]
NBA[,12] <- NBA[,12]/NBA[,5]
NBA[,14] <- NBA[,14]/NBA[,5]
NBA[,15] <- NBA[,15]/NBA[,5]
NBA[,16] <- NBA[,16]/NBA[,5]
NBA[,17] <- NBA[,17]/NBA[,5]
NBA[,18] <- NBA[,18]/NBA[,5]
NBA[,21] <- NBA[,21]/NBA[,5]
tablestat1 <- function(player){
player <- as.character(player)  
playerstats <- NBA[(NBA$Name == player),c(2,5,6,7,8) ]
names(playerstats)[3] <- "Minutes"
playerstats
}
tablestat2 <- function(player){
  player <- as.character(player)  
  playerstats <- NBA[(NBA$Name == player),c(2,9,10,11,12) ]
  playerstats
}
tablestat3 <- function(player){
  player <- as.character(player)  
  playerstats <- NBA[(NBA$Name == player),c(2,14,15,16,17,18,21) ]
  names(playerstats)[2] <- "Rebounds"
  names(playerstats)[7] <- "Points"
  playerstats
}
datap<- c(tablestat3("Rajon Rondo")[1,7],tablestat3("Rajon Rondo")[1,3],tablestat3("Rajon Rondo")[1,2],tablestat3("Rajon Rondo")[1,4],tablestat3("Rajon Rondo")[1,6],tablestat3("Rajon Rondo")[1,5])
datap<- as.numeric(datap)
df<- data.frame(cat =  factor(c("Points","Assists","Rebounds","Steals","Blocks","Turnover"),
       levels=c("Points","Assists","Rebounds","Steals","Blocks","Turnover")),stats=datap)
ggplot(data=df,aes(x = cat , y = stats,fill=cat)) + 
       geom_bar(stat = "identity") +  ylim(c(0,32.02)) + 
      ylab("")+xlab("")+scale_fill_hue(name="") + 
      ggtitle("General Stats")
```

---

## Improvements in future

Possibility to choose the season

Include team stats

Include average stats





